cardinalR: Generating interesting high-dimensional data structures

### 2025-05-07 paper-cardinalR.Rmd
Updated the mathematical formula
Updated notification

```{r main-tb-pdf, eval=knitr::is_latex_output()}
main_tb |> 
  kable(caption = "The main arguments for gen\\_multicluster().", format="latex", col.names = c("Argument", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

@@

The `gen_expbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ exponential branches in 2D, with optional noise dimensions to embed the structure in a \$p\$-dimensional space. These branches grow in opposite directions, producing a radiating curvilinear structure from a central region.

@@

$X\_1 \sim \mathcal{U}(-2, 2)$
$X\_2 = \exp(\pm s\_i \cdot X\_1) + \epsilon$,

@@

$s\_i \in [0.5, 2]$ is randomly sampled for each branch.

@@

$$
X\_j \sim \mathcal{N}(0, 0.1^2), \quad j = 3, \dots, p.
$$

@@

Each branch is a segment of a line with added jitter to simulate measurement noise. The branches differ in direction and location. Branches 1 and 2 are initialized with fixed slopes and intercepts. The Branch 1 is generated from $X\_1 \sim \mathcal{U}(-2, 8)\$, \$X\_2 = 0.5 \cdot X\_1 + \epsilon$. The Branch 2 is generated from $X\_1 \sim \mathcal{U}(-6, 2)\$, \$X\_2 = -0.5 \cdot X\_1 + \epsilon$. $\epsilon \sim \mathcal{U}(0, 0.5)$.

@@

$X\_1$ values span a short segment ($x\_{start}$ to $x\_{start} + 1$),

$X\_2 = s\_i \cdot (X\_1 - x\_{start}) + y\_{start} + \epsilon$ where $s\_i$ is a sampled slope from a filtered set and $\epsilon \sim \mathcal{U}(0, 0.2)$.

@@

$$
X\_j \sim \mathcal{N}(0, 0.05^2)\$ for \$j = 3, \dots, p.
$$

@@

Branch 1 is generated from $X\_1 \sim \mathcal{U}(0, 1)$, and $X\_2 = 0.1 \cdot X\_1 + 1 \cdot X\_1^2 + \epsilon$, where $\epsilon \sim \mathcal{U}(0, 0.05)$. This produces a gently upward-curving parabola in the right half-plane.

Branch 2 is generated from $X\_1 \sim \mathcal{U}(-1, 0)$, and $X\_2 = 0.1 \cdot X\_1 - 2 \cdot X\_1^2 + \epsilon$, where $\epsilon \sim \mathcal{U}(0, 0.05)$. This creates a steeper, leftward-facing curve in the left half-plane.

Branches $3$ to $k$** are added iteratively. Each new branch begins at a randomly chosen point within a restricted horizontal band: $X\_1 \in [-0.15, 0.15]$ (to ensure connectivity with earlier branches), spans a unit-length interval on $X\_1$:$X\_1 \in [x\_{\text{start}}, x\_{\text{start}} + 1]$, and has the structure:

@@

Here, $s\_i$ is a sampled scale factor (slope-like term) controlling curvature, and $(x\_{\text{start}}, y\_{\text{start}})$ is the starting point from the existing structure.

@@

To construct each branch, a unique or repeated pair of dimensions is selected from the $\binom{p}{2}$ possible $2\text{-}D$ combinations. If $k \leq \binom{p}{2}$, combinations are sampled without replacement. If $k > \binom{p}{2}$, additional pairs are sampled with replacement to reach $k$ total branches. Each selected pair $(x\_{i1}, x\_{i2})$ defines the $2\text{-}D$ plane for branch $i$.

For branch $i$, $n\_i$ points are generated, where $\sum\_{i=1}^k n\_i = n$. The structure follows:

@@

where $s\_i$ is a scale factor controlling the slope. When $k \leq \binom{p}{2}$, $s\_i = 1$. When sampling with replacement, $s\_i$ is drawn from the set ${1, 1.5, 2, \dots, 8}$ in increments of $0.5$.

@@

Let $\binom{p}{2}$ denote the number of unique $2\text{-}D$ subspace combinations. When $k \leq \binom{p}{2}$, $k$ distinct subspace pairs $(x\_{i1}, x\_{i2})$ are sampled without replacement. Otherwise, combinations are selected with replacement to reach $k$ total branches. For each branch $i = 1, \dots, k$, a random scale factor $s\_i$ is used to vary the curvature:

* If $k \leq \binom{p}{2}$: $s\_i = 1 $
* If $k > \binom{p}{2}$: $s\_i \sim \text{Uniform sample from } {1, 1.5, \dots, 8}$

Each branch contains $n\_i$ points such that $\sum\_{i=1}^k n\_i = n$, where the vector $(n\_1, \dots, n\_k)$ is randomly drawn using the helper function `gen_nsum()` to partition $n$.

Within its assigned subspace $(x\_{i1}, x\_{i2})$, branch $i$ is defined by:

@@

This forms a smooth downward-opening parabola in the plane defined by $(x\_{i1}, x\_{i2})$, with the degree of curvature controlled by $s\_i$.

@@

Each variable $X\_j$ represents integer grid coordinates (e.g., 1, 2, 3, ...) in dimension $j$:

$$
X\_j \in {1, \dots, k\_j},\text{ where }\prod\_{j=1}^p k\_j \approx n.
$$

@@

Points are sampled independently from $X\_1, X\_2, X\_3 \sim \mathcal{U}(-0.5, 0.5)$ of a cube of side length 1. This results in a uniform cloud of points filling the unit cube in $3\text{-}D$ space.

@@

$$
X\_j \sim \mathcal{N}(0, 0.05^2), \quad j = 4, \dots, p.
$$

@@

The initial points are drawn from a uniform cube centered at the origin: $X\_1, X\_2, X\_3 \sim \mathcal{U}(-0.5, 0.5)$.

@@

Remove if $\sqrt{X\_1^2 + X\_2^2 + X\_3^2} \leq 0.5$.

@@

$$
X\_j \sim \mathcal{N}(0, 0.05^2), \quad j = 4, \dots, p.
$$.

@@

$$
X\_1 = \left(1 + \frac{w}{2} \cos(\theta/2)\right) \cos(\theta),
$$
$$
X\_2 = \left(1 + \frac{w}{2} \cos(\theta/2)\right) \sin(\theta),
$$
$$
X\_3 = \frac{w}{2} \sin(\theta/2).
$$

@@

The curve is constructed by drawing uniformly spaced inputs and applying a second-degree polynomial transformation. Let $X\_1 \sim \mathcal{U}(\text{range}[1], \text{range}[2])$ be the independent variable. A raw polynomial basis of degree 2 is used to form $X\_2 = X\_1 - X\_1^2 + \varepsilon\_2$, where $\varepsilon\_2 \sim \mathcal{U}(0, 0.5)$.

@@

The shape is generated using a third-degree raw polynomial basis expansion. Let $X\_1 \sim \mathcal{U}(\text{range}[1], \text{range}[2])$ be the base input. A cubic transformation with vertical jitter is used to define $X\_2 = X\_1 + X\_1^2 - X\_1^3 + \varepsilon\_2\$, where \$\varepsilon\_2 \sim \mathcal{U}(0, 0.5)$.

@@

$$
X\_j \sim \mathcal{N}(0, 0.1^2)\$ for \$j = 3, \dots, p.
$$

@@

The `gen_pyrrect(n, p, h, l_vec, rt)` function generates a dataset of $n$ points forming a high-dimensional pyramid-like structure with a rectangular base. The pyramid is embedded in a $p$-dimensional space, with a tip at height zero and base at height $h$. The shape tapers linearly from the base dimensions $(l\_x, l\_y)$ to a smaller rectangular section with side lengths $(2r\_t, 2r\_t)$ at the tip. Points are distributed more densely near the tip, simulating a natural skew toward smaller height values.

Let $x\_1, x\_2, \dots, x\_p$ denote the coordinates of the generated points. The final dimension $x\_p$ encodes the height of each point and is drawn from an exponential distribution capped at $h$:

@@

Let $r\_x(z)$ and $r\_y(z)$ denote the half-widths of the rectangular cross-section at height $z$:

@@

For $p > 3$, the remaining $p - 4$ dimensions (i.e., $x\_4$ to $x\_{p-1}$) are tapered toward zero with decreasing height and generated as:

@@

Let $x\_1, x\_2, \dots, x\_p$ denote the coordinates of the generated points. The final coordinate \$x\_p\$ encodes the height $z$ of each point and is drawn from an exponential distribution capped at $h$:

@@

Let $x\_1, x\_2, \dots, x\_p$ denote the coordinates of the generated points. The final coordinate $x\_p$ encodes the height $z$ of each point and is drawn from an exponential distribution capped at $h$:

@@

Let $x\_1, x\_2, \dots, x\_p$ denote the coordinates of the generated points. The final coordinate $x\_p$ encodes the height $z$:

@@

The radius at height $z$ scales linearly from zero (tip) to the base radius $r\_b$:

@@

Each point is placed within a regular hexagon in the plane $(x\_1, x\_2)$, using a randomly chosen hexagon sector angle $\theta \in \{0, \frac{\pi}{3}, \frac{2\pi}{3}, \pi, \frac{4\pi}{3}, \frac{5\pi}{3}\}$ and a uniformly random radial scaling factor:

@@

Let $x\_1, x\_2, \dots, x\_p$ denote the coordinates of the generated points. The generation process begins with an initial point $T\_0 \in [0, 1]^p$ drawn from a uniform distribution:

@@

Let $C\_1, C\_2, \dots, C\_{p+1}$ denote the corner vertices of a $p$-dimensional simplex. At each iteration $i = 1, \dots, n$, a new point is computed by taking the midpoint between the previous point $T\_{i-1}$ and a randomly selected vertex $C\_k$:

@@

$$
X\_j \sim \mathcal{N}(0, 0.05^2), \quad j = 5, \dots, p.
$$

@@

* $X_1 \sim \mathcal{U}(0.1, 2)$: base variable (avoids zero to prevent division errors).
* $X_3 \sim \mathcal{U}(0.1, 0.8)$: independent auxiliary variable.
* $X_2 = \frac{hc}{X_1} + \text{non\_fac} \cdot \sin(X_1)$: non-linear combination of **hyperbolic and sinusoidal transformations**, creating sharp curvature and oscillation.
* $X_4 = \cos(\pi \cdot X_1) + \varepsilon$, with $\varepsilon \sim \mathcal{U}(-0.1, 0.1)$: additional nonlinear variation based on cosine, simulating more subtle periodic structure.

>>>>

```{r main-tb-pdf, eval=knitr::is_latex_output()}
main_tb |> 
  kable(caption = "The main arguments for gen_multicluster().", format="latex", col.names = c("Argument", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

@@

The `gen_expbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ exponential branches in $2\text{-}D$, with optional noise dimensions to embed the structure in a $p$-dimensional space. These branches grow in opposite directions, producing a radiating curvilinear structure from a central region.

@@

$X_1 \sim \mathcal{U}(-2, 2)$
$X_2 = \exp(\pm s_i \cdot X_1) + \epsilon$,

@@

$s_i \in [0.5, 2]$ is randomly sampled for each branch.

@@

$$
X_j \sim \mathcal{N}(0, 0.1^2), \quad j = 3, \dots, p.
$$

@@

Each branch is a segment of a line with added jitter to simulate measurement noise. The branches differ in direction and location. Branches 1 and 2 are initialized with fixed slopes and intercepts. The Branch 1 is generated from $X_1 \sim \mathcal{U}(-2, 8)\$, \$X_2 = 0.5 \cdot X_1 + \epsilon$. The Branch 2 is generated from $X_1 \sim \mathcal{U}(-6, 2)\$, \$X_2 = -0.5 \cdot X_1 + \epsilon$. $\epsilon \sim \mathcal{U}(0, 0.5)$.

@@

$X_1$ values span a short segment ($x_{start}$ to $x_{start} + 1$),

$X_2 = s_i \cdot (X_1 - x_{start}) + y_{start} + \epsilon$ where $s_i$ is a sampled slope from a filtered set and $\epsilon \sim \mathcal{U}(0, 0.2)$.

@@

$$
X_j \sim \mathcal{N}(0, 0.05^2)\text{ for }j = 3, \dots, p.
$$

@@

Branch 1 is generated from $X_1 \sim \mathcal{U}(0, 1)$, and $X_2 = 0.1 \cdot X_1 + 1 \cdot X_1^2 + \epsilon$, where $\epsilon \sim \mathcal{U}(0, 0.05)$. This produces a gently upward-curving parabola in the right half-plane.

Branch 2 is generated from $X_{1} \sim \mathcal{U}(-1, 0)$, and $X_2 = 0.1 \cdot X_{1} - 2 \cdot X_{1}^2 + \epsilon$, where $\epsilon \sim \mathcal{U}(0, 0.05)$. This creates a steeper, leftward-facing curve in the left half-plane.

Branches $3$ to $k$** are added iteratively. Each new branch begins at a randomly chosen point within a restricted horizontal band: $X_{1} \in [-0.15, 0.15]$ (to ensure connectivity with earlier branches), spans a unit-length interval on $X_{1}$:$X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]$, and has the structure:

@@

Here, $s_i$ is a sampled scale factor (slope-like term) controlling curvature, and $(x_{\text{start}}, y_{\text{start}})$ is the starting point from the existing structure.

@@

To construct each branch, a unique or repeated pair of dimensions is selected from the $\binom{p}{2}$ possible $2\text{-}D$ combinations. If $k \leq \binom{p}{2}$, combinations are sampled without replacement. If $k > \binom{p}{2}$, additional pairs are sampled with replacement to reach $k$ total branches. Each selected pair $(x_{i1}, x_{i2})$ defines the $2\text{-}D$ plane for branch $i$.

For branch $i$, $n_i$ points are generated, where $\sum_{i=1}^k n_i = n$. The structure follows:

@@

where $s_i$ is a scale factor controlling the slope. When $k \leq \binom{p}{2}$, $s_i = 1$. When sampling with replacement, $s_i$ is drawn from the set ${1, 1.5, 2, \dots, 8}$ in increments of $0.5$.

@@

Let $\binom{p}{2}$ denote the number of unique $2\text{-}D$ subspace combinations. When $k \leq \binom{p}{2}$, $k$ distinct subspace pairs $(x_{i1}, x_{i2})$ are sampled without replacement. Otherwise, combinations are selected with replacement to reach $k$ total branches. For each branch $i = 1, \dots, k$, a random scale factor $s_i$ is used to vary the curvature:

* If $k \leq \binom{p}{2}$: $s_i = 1 $
* If $k > \binom{p}{2}$: $s_i \sim \text{Uniform sample from } {1, 1.5, \dots, 8}$

Each branch contains $n_i$ points such that $\sum_{i=1}^k n_i = n$, where the vector $(n_1, \dots, n_k)$ is randomly drawn using the helper function `gen_nsum()` to partition $n$.

Within its assigned subspace $(x_{i1}, x_{i2})$, branch $i$ is defined by:

@@

This forms a smooth downward-opening parabola in the plane defined by $(x_{i1}, x_{i2})$, with the degree of curvature controlled by $s_i$.

@@

Each variable $X_j$ represents integer grid coordinates (e.g., 1, 2, 3, ...) in dimension $j$:

$$
X_j \in {1, \dots, k_j},\text{ where }\prod_{j=1}^p k_j \approx n.
$$

@@

Points are sampled independently from $X_1, X_2, X_3 \sim \mathcal{U}(-0.5, 0.5)$ of a cube of side length 1. This results in a uniform cloud of points filling the unit cube in $3\text{-}D$ space.

@@

$$
X_j \sim \mathcal{N}(0, 0.05^2), \quad j = 4, \dots, p.
$$

@@

The initial points are drawn from a uniform cube centered at the origin: $X_1, X_2, X_3 \sim \mathcal{U}(-0.5, 0.5)$.

@@

Remove if $\sqrt{X_1^2 + X_2^2 + X_3^2} \leq 0.5$.

@@

$$
X_j \sim \mathcal{N}(0, 0.05^2), \quad j = 4, \dots, p.
$$.

@@

$$
X_1 = \left(1 + \frac{w}{2} \cos(\theta/2)\right) \cos(\theta),
$$
$$
X_2 = \left(1 + \frac{w}{2} \cos(\theta/2)\right) \sin(\theta),
$$
$$
X_3 = \frac{w}{2} \sin(\theta/2).
$$

@@

The curve is constructed by drawing uniformly spaced inputs and applying a second-degree polynomial transformation. Let $X_1 \sim \mathcal{U}(\text{range}[1], \text{range}[2])$ be the independent variable. A raw polynomial basis of degree 2 is used to form $X_2 = X_1 - X_1^2 + \varepsilon_2$, where $\varepsilon_2 \sim \mathcal{U}(0, 0.5)$.

@@

The shape is generated using a third-degree raw polynomial basis expansion. Let $X_1 \sim \mathcal{U}(\text{range}[1], \text{range}[2])$ be the base input. A cubic transformation with vertical jitter is used to define $X_2 = X_1 + X_1^2 - X_1^3 + \varepsilon_2\$, where \$\varepsilon_2 \sim \mathcal{U}(0, 0.5)$.

@@

$$
X_j \sim \mathcal{N}(0, 0.1^2)\$ for \$j = 3, \dots, p.
$$

@@

The `gen_pyrrect(n, p, h, l_vec, rt)` function generates a dataset of $n$ points forming a high-dimensional pyramid-like structure with a rectangular base. The pyramid is embedded in a $p$-dimensional space, with a tip at height zero and base at height $h$. The shape tapers linearly from the base dimensions $(l_x, l_y)$ to a smaller rectangular section with side lengths $(2r_t, 2r_t)$ at the tip. Points are distributed more densely near the tip, simulating a natural skew toward smaller height values.

Let $x_1, x_2, \dots, x_p$ denote the coordinates of the generated points. The final dimension $x_p$ encodes the height of each point and is drawn from an exponential distribution capped at $h$:

@@

Let $r_x(z)$ and $r_y(z)$ denote the half-widths of the rectangular cross-section at height $z$:

@@

For $p > 3$, the remaining $p - 4$ dimensions (i.e., $x_4$ to $x_{p-1}$) are tapered toward zero with decreasing height and generated as:

@@

Let $x_1, x_2, \dots, x_p$ denote the coordinates of the generated points. The final coordinate \$x_p\$ encodes the height $z$ of each point and is drawn from an exponential distribution capped at $h$:

@@

Let $x_1, x_2, \dots, x_p$ denote the coordinates of the generated points. The final coordinate $x_p$ encodes the height $z$ of each point and is drawn from an exponential distribution capped at $h$:

@@

Let $x_1, x_2, \dots, x_p$ denote the coordinates of the generated points. The final coordinate $x_p$ encodes the height $z$:

@@

The radius at height $z$ scales linearly from zero (tip) to the base radius $r_b$:

@@

Each point is placed within a regular hexagon in the plane $(x_1, x_2)$, using a randomly chosen hexagon sector angle $\theta \in \{0, \frac{\pi}{3}, \frac{2\pi}{3}, \pi, \frac{4\pi}{3}, \frac{5\pi}{3}\}$ and a uniformly random radial scaling factor:

@@

Let $x_1, x_2, \dots, x_p$ denote the coordinates of the generated points. The generation process begins with an initial point $T_0 \in [0, 1]^p$ drawn from a uniform distribution:

@@

Let $C_1, C_2, \dots, C_{p+1}$ denote the corner vertices of a $p$-dimensional simplex. At each iteration $i = 1, \dots, n$, a new point is computed by taking the midpoint between the previous point $T_{i-1}$ and a randomly selected vertex $C_k$:

@@

@@

* $X_1 \sim \mathcal{U}(0.1, 2)$: base variable (avoids zero to prevent division errors).
* $X_3 \sim \mathcal{U}(0.1, 0.8)$: independent auxiliary variable.
* $X_2 = \frac{hc}{X_1} + \text{non_fac} \cdot \sin(X_1)$: non-linear combination of **hyperbolic and sinusoidal transformations**, creating sharp curvature and oscillation.
* $X_4 = \cos(\pi \cdot X_1) + \varepsilon$, with $\varepsilon \sim \mathcal{U}(-0.1, 0.1)$: additional nonlinear variation based on cosine, simulating more subtle periodic structure.
